Estimating duration in the presence of misclassification

### 2024-01-25 transmission.tex
almost all chapter text written

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining priors had weakly informative priors.

@@

For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown\todo{consider showing prior/posterior plots}) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
The decreases in $\beta_w$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
A multiple variant model might be able to capture this rise more accurately.
Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
Then, each variant would be explicitly included in the model to fit to this data.

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
This implies that the data would allow identification of a more flexible age-based model.
Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $\beta_c$.
More parameters could capture the age distribution more accurately.
There are various ways to increase the model's flexibility, including introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
It is plausible that these distributions changed over time in a way not captured by $\matr{\kappa}$, for example due to the introduction and relaxation of control measures within schools.
$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.

The weekly patterns in incidence are another limitation of the current parameterisation.
The random walk and contact matrices both vary only on the weekly timescale.
A daily random walk could provide more flexibility here.
However, this would greatly increase the number of parameters.
An alternative would be a smoothly-varying function or an approximation of a random walk such as that proposed by \textcite{ghoshApproximate}.

@@

Comparing the regional incidence estimates between the two models in this chapter.

Example of bias---variance trade-off.

Demonstrated by fitting to the North East.
Quality of fit in SEIR model is variable between different age groups.
This is because the model is not flexible enough to capture the true age distribution of infections.

\begin{itemize}
    \item Results here disagree with previous chapter's results.
    \item The model in previous chapter is more flexible than here, showing the limitations discussed above. Previous chapter fits better.
    \item The previous chapter's modelling benefits from being able to use flexible approximate inference methods that scale to many parameters, including random effects. Here, I am limited by the inference procedure. MCMC was unable to scale to the full model used in the previous chapter, which has a computationally cheaper likelihood as it is a GLM, not requiring solving an ODE.
    \item Shows that there is a lot of information in the data which could be exploited by a suitably flexible mechanistic model. Such a model would be useful for policy-decisions because it could infer where transmission is occurring (\eg in schools or workplaces) and how this changes over time.
    \item The narrower credible intervals in this chapter are a realisation of the standard bias-variance tradeoff. By making stronger assumptions regarding how transmission occurs, the variance is reduced. However, if these assumptions are incorrect (which is almost certainly the case), bias occurs.
\end{itemize}

\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}
\section{Conclusion}

>>>>

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining parameters had weakly informative priors.

@@

For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.

The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or interventions not captured by $\matr{\kappa}$ such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
Local measures are not well captured by $\matr{\kappa}$ because the matrix averages behaviour across the whole of England.
Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.

The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite[e.g.][]{daviesEstimated}.
A multiple variant model might be able to capture this rise more accurately.
Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
Then, each variant would be explicitly included in the model to fit to this data.

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
This implies that the data would allow identification of a more flexible age-based model.

Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $q_c$.
Increasing the number of parameters in $\matr{\beta}$ would be the simplest approach to increase the model's flexibility.

The only time-changing part of $\matr{\beta}$ that varies between ages is $\matr{\kappa}$.
It is plausible that this would not capture all behavioural changes.
For example, the introduction and relaxation of control measures within schools.
$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.

The weekly patterns in incidence are a limitation of the current parameterisation.
The random walk and contact matrices both vary only on the weekly timescale.
A daily random walk could provide more flexibility here.
However, this would greatly increase the number of parameters.
An alternative would be a smoothly-varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally the estimates agree, with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
However, even here, the credible intervals overlap except in London.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

These patterns can be interpreted as an instance of the common bias---variance trade-off.
The SEIR model imposes stronger assumptions about the epidemic dynamics.
The assumptions reduces the variance of the estimates, including making them smoother, but if the assumptions are incorrect, bias is introduced.
This trade-off based on the strength of assumptions is well-known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
The SEIR model can be viewed as a backcalulation method where the transmisson part of the model is a smoothing method.

The backcalculation method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
The SEIR model is limited by the inference procedure and the requirement to sovle an ODE to evaluate the likelihood.
To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}

\section{Conclusion}
